---
title: "Line profiles for sparse version SuSiE"
author: "Kaiqian Zhang"
date: "9/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
devtools::install_github("stephenslab/susieR")
```

```{r, warning=FALSE}
library(Matrix)
library(susieR)
library(profvis)
```

## Goal
In this vignette, we provide line profiles for revised version SuSiE, which allows for a sparse matrix structure. We compare speed performance when the form of the matrix `X` is dense and sparse. 

## Results

In this minimal example, we observe that given a large sparse matrix, if it is in the dense form, the speed is around `40%` slower than that in a sparse form.  

## Data
We randomly simulate a `n=1000` by `p=1000` dense matrix and a sparse matrix at sparsity $99\%$, i.e. $99\%$ entries are zeros.
```{r}
create_sparsity_mat = function(sparsity, n, p){
  set.seed(1)
  nonzero = round(n*p*(1-sparsity))
  nonzero.idx = sample(n*p, nonzero)
  mat = numeric(n*p)
  mat[nonzero.idx] = 1
  mat = matrix(mat, nrow=n, ncol=p)
  return(mat)     
}
```

```{r}
n = 1000
p = 1000
beta = rep(0,p)
beta[1]    = 10 
beta[300]  = 10
beta[400]  = 10
beta[1000] = 10
X.dense = create_sparsity_mat(0.99,n,p)
X.sparse = as(X.dense,'dgCMatrix')
y = c(X.dense %*% beta + rnorm(n))
```

## `X` in a dense form
```{r}
profvis({
  susie.dense = susie(X.dense, y)
})
```

## `X` in a sparse form
```{r}
profvis({
  susie.sparse = susie(X.sparse, y)
})
```

## Further step
We encourage people who are insterested in improving SuSiE can get insights from those line profiles provided. 
