---
title: "Intro"
author: "Matthew Stephens"
date: "April 11, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is just a quick document to show some examples of `susie` in action.


# Simple simulation

Here we simulate data with four non-zero effects.
```{r}
library(susieR)
set.seed(1)
n = 1000
p = 1000
beta = rep(0,p)
beta[1] = 1
beta[2] = 1
beta[3] = 1
beta[4] = 1
X = matrix(rnorm(n*p),nrow=n,ncol=p)
y = X %*% beta + rnorm(n)

res =susie(X,y,L=10)
plot(coef(res))
plot(y,predict(res))
```


# Trend filtering

This is an example of using susie to do trend filtering.

```{r}
set.seed(1)
n=1000
delta = matrix(1,nrow=n,ncol=n)
for(i in 1:(n-1)){
  for(j in (i+1):n){
    delta[i,j] = -1
  }
}

beta = c(rep(0,100),rep(1,100),rep(3,100),rep(-2,100),rep(0,600))
y = beta + rnorm(n)
s = susie(delta,y,sigma2=1,L=10)
```

Plot results: the truth is green, and susie estimate is red.
```{r}
plot(y)
lines(predict(s),col=2,lwd=3)
lines(beta,col=3,lwd=3)
```


Try something harder where the beta increases linearly:
```{r}
set.seed(1)
beta = seq(0,1,length=1000)
y = beta + rnorm(n)
s = susie(delta,y,sigma2=1,L=10)
plot(y)
lines(predict(s),col=2,lwd=3)
lines(beta,col=3,lwd=3)
```

Compare with the lasso based solution
```{r}
library("genlasso")
y.tf = trendfilter(y,ord=1)
y.tf.cv = cv.trendfilter(y.tf)
plot(y)
lines(predict(s),col=2,lwd=3)
lines(beta,col=3,lwd=3)
lines(y.tf$fit[,which(y.tf$lambda==y.tf.cv$lambda.min)],col=4,lwd=3)
#plot(y.tf,lambda=y.tf.cv$lambda.min,col=2)
```

What happens if we have trend plus sudden change.

```{r}
beta = beta + c(rep(0,500),rep(2,500))
y = beta + rnorm(n)
s = susie(delta,y,sigma2=1,L=10)
plot(y)
lines(predict(s),col=2,lwd=3)
lines(beta,col=3,lwd=3)

# trend
y.tf = trendfilter(y,ord=1)
y.tf.cv = cv.trendfilter(y.tf)
lines(y.tf$fit[,which(y.tf$lambda==y.tf.cv$lambda.min)],col=4,lwd=3)
```


