---
title: "SuSiE with `L0Learn` initialization example"
author: "Kaiqian Zhang"
date: "July 18, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#devtools::install_github("stephenslab/susieR", force = TRUE)
#devtools::install_github("hazimehh/L0Learn", force = TRUE)
```

`susieR` allows for a customized initialization. In this vignette we deomonstrate how to use [`L0Learn`](https://github.com/hazimehh/L0Learn) fit to initialize susieR.

```{r, warning=FALSE}
library(susieR)
library(L0Learn)
```

## Simulate data

First, we simulate data from the minimal example.

```{r}
set.seed(1)
n = 1000
p = 1000
beta = rep(0,p)
beta[1] = 1
beta[2] = 1
beta[3] = 300
beta[4] = 400
X = matrix(rnorm(n*p),nrow=n,ncol=p)
y = X %*% beta + rnorm(n)
```

## Fit L0Learn 

We start with fitting a L0-regularized model to the simulated data. The L0Learn model finds 115 effects, which will be used to initialize susie.

```{r}
set.seed(1)
L0fit = L0Learn.cvfit(X, y, penalty = "L0")
```

```{r}
lambdaIndex = which.min(L0fit$cvMeans[[1]]) # find the optimal lambda, which has minimum cv error
L0coef = as.numeric(coef(L0fit$fit, lambda = L0fit$fit$lambda[[1]][lambdaIndex]))
effect.beta = L0coef[which(L0coef!=0)][-1]
effect.index = (which(L0coef!=0)-1)[-1] 
length(effect.beta)
effect.beta[1:10]
effect.index[1:10]
```

## Build an initialization object

We create an initialization using `susie_set_init` function

```{r}
set.seed(1)
s.init = susie_set_init(effect.index, effect.beta, num_variables=p, prior_variance=0.2, residual_variance=as.numeric(var(y)))
```

## Run susieR with initialization

Now, we use `effect.index` and `effect.beta` obtained from L0Learn fit to create an `s.init` object. We then run susie with this initialization.  

```{r}
susieL0.fit = susie(X,y,s_init=s.init)
susie.CS.set = susie_get_CS(susieL0.fit, X=X)
susie.CS.set$cs
```

## References

Hussein Hazimeh and Rahul Mazumder. (2018). Fast Best Subset Selection: Coordinate Descent and Local Combinatorial Optimization Algorithms