---
title: "SuSiE with a genotype slider"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SuSiE with a genotype slider}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The root package in this repository is `susieRSlidePrior`. Its default `susie()`
fits `x + delta * I(x == 1)` using a fixed finite prior on `delta`, separately
for each candidate SNP and single-effect component. The default grid is
`seq(-1, 1, length.out = 17)` with equal probabilities. `delta = -1` is recessive,
`delta = 0` additive, and `delta = 1` dominant.

```{r}
library(susieRSlidePrior)
set.seed(341)
X <- matrix(rbinom(600 * 20, 2, 0.3), 600, 20)
storage.mode(X) <- "double"
colnames(X) <- paste0("SNP", seq_len(ncol(X)))
y <- 0.5 + 1.5 * (X[, 3] == 2) - 0.8 * (X[, 14] >= 1) +
  rnorm(nrow(X), sd = 0.4)
fit <- susie(X, y, L = 2, min_obs = 5)
fit$delta_cs$summary
fit$delta_cs$delta
fit$delta_prior
fit$delta_weights

# The supplied probabilities stay fixed throughout this fit.
w <- rep(1, 17)
w[9] <- 8
fit_weighted <- susie(X, y, L = 2, delta_prior = w / sum(w))
stopifnot(isTRUE(all.equal(fit_weighted$delta_prior, w / sum(w))))
```

`delta_cs$summary` lists member SNPs in each reported credible set (CS).
`delta_cs$delta` is a CS-by-SNP matrix: rows follow `sets$cs`, mapped to
component rows by `sets$cs_index`; columns include all input SNPs in their
original order, excluding an explicit null column. Each entry is that
component's posterior mean delta for the candidate SNP, even outside the CS.
The full component-by-SNP matrix remains available as `fit$delta`.

Supply original `0/1/2` genotypes. Centering and standardization are performed
internally, dividing both genotype and heterozygote terms by the original
additive genotype SD. The Gaussian coefficient prior uses those internal
units, as in SuSiE. Output deltas keep their original interpretation.

```{r}
stopifnot(identical(dim(fit$delta), dim(fit$alpha)))
head(predict(fit, newx = X))
coef(fit)[c(1, 4, 15), , drop = FALSE]
```

`min_obs = 5` fixes delta at **zero** if any genotype class has fewer than
five retained observations, including an absent class. `coef()` returns
additive and heterozygote coefficients because both are needed for prediction.

The original additive fitter is available explicitly:

```{r}
additive <- susie_additive(X, y, L = 2)
fixed_zero <- susie(X, y, L = 2, delta = 0)
max(abs(additive$fitted - fixed_zero$fitted))
```

The slider integrates beta analytically and sums over the fixed delta prior.
`alpha_delta[l,j,k]` stores the joint posterior mass for each SNP and slider
value; summing over k gives the usual SNP-level `alpha`. `mu_delta` retains
E(beta*delta), so predictions account for their posterior dependence.
`delta_prior_counts` excludes forced SNPs, null columns and inactive components
and can supply a later external coding-prior learning workflow. The fitter
does not perform that M-step. `estimate_prior_method="EM"` concerns only the
Gaussian beta-prior variance. Supply `delta_prior=NULL` to run the legacy
continuous plug-in slider instead. CS purity is computed from posterior-mean
slider codings as a representative-coding diagnostic. The model currently
requires complete hard-call genotypes. The inherited summary-statistic
interfaces remain additive and do not estimate sliders.
