---
title: "SuSiE with a genotype slider"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SuSiE with a genotype slider}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The root package in this repository is `susieSlide`. Its default `susie()`
fits `x + delta * I(x == 1)`, with `delta` in `[-1, 1]`, separately for each
candidate SNP and single-effect component. `delta = -1` is recessive,
`delta = 0` additive, and `delta = 1` dominant.

```{r}
library(susieSlide)
set.seed(341)
X <- matrix(rbinom(600 * 20, 2, 0.3), 600, 20)
storage.mode(X) <- "double"
colnames(X) <- paste0("SNP", seq_len(ncol(X)))
y <- 0.5 + 1.5 * (X[, 3] == 2) - 0.8 * (X[, 14] >= 1) +
  rnorm(nrow(X), sd = 0.4)
fit <- susie(X, y, L = 2, min_obs = 5)
fit$delta_cs
```

Supply original `0/1/2` genotypes. Centering and standardization are performed
internally, dividing both genotype and heterozygote terms by the original
additive genotype SD. The Gaussian coefficient prior uses those internal
units, as in SuSiE. Output deltas keep their original interpretation.

```{r}
stopifnot(identical(dim(fit$delta), dim(fit$alpha)))
head(predict(fit, newx = X))
coef(fit)[c(1, 4, 15), , drop = FALSE]
```

`min_obs = 5` fixes delta at **zero** if any genotype class has fewer than
five retained observations, including an absent class. `coef()` returns
additive and heterozygote coefficients because both are needed for prediction.

The original additive fitter is available explicitly:

```{r}
additive <- susie_additive(X, y, L = 2)
fixed_zero <- susie(X, y, L = 2, delta = 0)
max(abs(additive$fitted - fixed_zero$fitted))
```

The slider integrates beta analytically and estimates delta by maximizing
marginal likelihood. Evidence scores, PIPs and credible sets condition on
fitted delta. Null simulations show additional fitting from optimizing
delta; genome-wide calibration is not established. The model currently
requires complete hard-call genotypes. The inherited summary-statistic
interfaces remain additive and do not estimate sliders.
