---
title: "A minimal example on revised SuSiE"
author: "Kaiqian Zhang"
date: "9/10/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE}
library(Matrix)
library(ggplot2)
library(microbenchmark)
library(profvis)
```

## Goal
In this vignette, we want to compare a revised version SuSiE, which allows for sparse matrix multiplication, with the original one. We claim that the revised version reduces computation time by a large extent and saves more memory. Since we have verified correctness by unit tests, we will only consider speed comparisons here. 

## Results

In this minimal example, we have shown that the revised version of SuSiE can reduce computation time by around $40\%$, even if the input matrix is a dense matrix, and further reduce computation time $5\%$ more given the input matrix is a sparse matrix. It is also obvious that the revised version takes less memory. We provide line profiling below so that people who are interested could improve the work as well.  

## Data
We randomly simulate a dense matrix and a sparse matrix at sparsity $99\%$, i.e. $99\%$ entries are zeros.
```{r}
create_sparsity_mat = function(sparsity, n, p){
  nonzero = round(n*p*(1-sparsity))
  nonzero.idx = sample(n*p, nonzero)
  mat = numeric(n*p)
  mat[nonzero.idx] = 1
  mat = matrix(mat, nrow=n, ncol=p)
  return(mat)     
}
```

```{r}
n = 1000
p = 1000
beta = rep(0,p)
beta[1]    = 10 
beta[300]  = 10
beta[400]  = 10
beta[1000] = 10
set.seed(1)
X.dense = create_sparsity_mat(0.99,n,p)
X.sparse = as(X.dense,'dgCMatrix')
y = c(X.dense %*% beta + rnorm(n))
```

## Original version SuSiE
```{r}
devtools::install_github("stephenslab/susieR")
library(susieR)
```

```{r}
profvis({
  susie.original = susie(X.dense, y)
})
```

## Revised version SuSiE
```{r}
devtools::install_local("~/Desktop/M/M-github-repos/susieR-sparse-v3")
library(susieR)
```

```{r}
profvis({
  susie.revised.dense = susie(X.dense, y)
})
```

```{r}
profvis({
  susie.revised.sparse = susie(X.sparse, y)
})
```

