---
title: "SuSiE with stumps"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries:

```{r}
library(susieR)
```


# Introduction

My goal here is to test the idea of using stumps to do (non-)linear regression with SuSiE.

Let $y$ be an $n$ vector $X$ be an $n \times p$ matrix of covariates,
and let $x_j$ denote the $j$th column of $X$.

The idea is that for each covariate vector $x_j$, we can let $y$ be non-linearly dependent on $x_j$ by using non-parametric regression - e.g. using the ranks of the elements of $x_j$ as the "time" axis, and using a changepoint (piecewise constant) basis. (Alternatively we could use a wavelet basis,  but those are not "stumps".)

In the case $p=1$ we have effectively already implemented this (`susie_tf`). The idea is to generalize this beyond $p=1$.

# Simulate with p=1

Here the results should match trend filtering....
```{r}
set.seed(1)
n=100
X = cbind(rnorm(n))
y = (X[,1]>1)  + rnorm(n,0,0.1)
s = susie_stumps_old(X[,1,drop=FALSE],y,standardize=FALSE)

plot(rank(X[,1]),y)
o = order(X[,1])
lines(sort(rank(X[,1])),(s$Xr+mean(y))[o])

s2 = susie_trendfilter(y[o],0,use_mad=FALSE,standardize = FALSE)
#plot(sort(rank(X[,1])),y[o])
lines(sort(rank(X[,1])),s2$Xr+mean(y),col=2)

s$Xr[o] - s2$Xr

s$sigma2 - s2$sigma2
```

Here we try the new list-based approach...
```{r}
s3 = susie_stumps(X,y,include_linear = FALSE)
s3$Xr-s$Xr
s3$sigma2 - s$sigma2
```


# Case p=2

Simulate some data to illustrate the idea
```{r}
X = cbind(rnorm(n),rnorm(n))
y = (X[,1]>0) + (X[,2]>1) + rnorm(n,0,0.1)
plot(X[,1],y)
plot(X[,2],y)
```

```{r}
Xord = apply(X,2,order) #the idea is we want to order y by these two different orderings
plot(y[Xord[,1]])
plot(y[Xord[,2]])
```

```{r}
s = susie_stumps_old(X,y,L=2,standardize=FALSE)
plot(s$fitted,y)
abline(a=0,b=1)
```

Here we try the new list-based approach...
```{r}
s2 = susie_stumps(X,y,L=2)
s2$Xr-s$Xr
plot(s2$fitted,y)
plot(s2$fitted, s$fitted)
s$V
s2$V
s$elbo
s2$elbo
```


