---
title: "SuSiE with summary statistics"
author: "Yuxin Zou"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SuSiE with summary statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,comment = "#",fig.width = 5,
                      fig.height = 3,fig.align = "center",
                      fig.cap = "&nbsp;",dpi = 120)
```

This vignette demonstrates how to use `susieR` with summary statistics in the context of genetic fine-mapping. We use the same simulated data as in [fine mapping vignette](finemapping.html). The simulated data is expression level of a gene ($y$) in $N \approx 600$ individuals. We want to identify with the genotype matrix $X_{N\times P}$ ($P=1000$) the genetic variables that causes changes in expression level.

The simulated data-set is [availablehere][N3finemapping], as part of the `susieR` package. It is simulated to have exactly 3 non-zero effects.

```{r}
library(susieR)
set.seed(1)
```

## The data-set

```{r}
finemap <-
  readRDS(system.file("datafiles","N3finemapping.rds",package = "susieR"))
```

Notice that we've simulated 2 sets of $Y$ as 2 simulation replicates. Here we'll focus on the first data-set.

```{r}
dim(finemap$data$Y)
```

Here are the 3 "true" signals in the first data-set:

```{r}
b <- finemap$data$true_coef[,1]
plot(b, pch=16, ylab='effect size')
```

```{r}
which(b != 0)
```

So the underlying causal variables are 403, 653 and 773.

## Simple regression summary statistics

The data-set additionally provides summary statistics for fitting univariate simple regression variable by variable. The results are $\hat{\beta}$ and $SE(\hat{\beta})$ from which the p values abd $z$-scores can be derived. Again we focus only on results from the first data-set:

```{r}
p_values = 2*pt(-abs(finemap$sumstats[1,,1]/finemap$sumstats[2,,1]), df = nrow(finemap$data$X)-2)
z_scores = abs(qnorm(p_values/2)) * sign(finemap$sumstats[1,,1])
susie_plot(z_scores, y = "z", b=b)
```

The correlation matrix is
```{r}
R <- cor(finemap$data$X)
```

## Fine-mapping with `susieR` with the knowledge of sample size n

For starters, we assume there are at most 5 causal variables, i.e. set `L=5`, although SuSiE is robust to the choice of `L`.

We set SuSiE prior variance to 0.1. 

### Using bhat and shat

With the knowledge of variance of y, the `susieR` function call gives coefficients in the original X, y scale is:

```{r}
fitted_bhat <- susie_bhat(bhat = finemap$sumstats[1,,1], shat = finemap$sumstats[2,,1], 
                          R = R, n = nrow(finemap$data$X), 
                          var_y = var(finemap$data$Y[,1]),
                          L = 5, 
                          scaled_prior_variance = 0.1, 
                          estimate_prior_variance = FALSE, 
                          standardize = TRUE)
```

Without the knowledge of variance of y, `susieR` gives the coefficients in standaridized X, y scale.

```{r}
fitted_bhat_standardize <- susie_bhat(bhat = finemap$sumstats[1,,1], shat = finemap$sumstats[2,,1], 
                                      R = R, n = nrow(finemap$data$X), 
                                      L = 5, 
                                      scaled_prior_variance = 0.1, 
                                      estimate_prior_variance = FALSE)
```

Here is the summary of the input and the output coefficient scale:

  known var(y)    |  coefficient scale
------------------|-------------------
  yes             |  same as bhat, shat
------------------|-------------------
  no              |  standardized X, y

Using `summary` function, we can see the posterior inclusion probability (PIP) for each variable, and the 95% confidence sets. Here, we only show the 95% confidence sets.

```{r}
summary(fitted_bhat)$cs
```

The 3 causal signals have been captured by the 3 CS reported here. The 3rd CS contains many variables, including the true causal variable 403.

We can also plot the PIP
```{r}
susie_plot(fitted_bhat, y="PIP", b=b)
```

The true causal variables are colored red. The 95% CS identified are circled in different colors.

### Using z scores

Using z scores, `susieR` gives the coefficients in standaridized X, y scale.

```{r}
fitted_z_n <- susie_bhat(bhat = z_scores, shat = 1, 
                         R = R, n = nrow(finemap$data$X), 
                         L = 5, 
                         scaled_prior_variance = 0.1, 
                         estimate_prior_variance = FALSE)
```

We can check the 95% confidence set
```{r}
summary(fitted_z_n)$cs
```

and the PIP plot
```{r}
susie_plot(fitted_bhat, y="PIP", b=b)
```

## Fine-mapping with `susieR` without the knowledge of sample size n

Without the sample size n, the inference on effects are more conservative.

Again, we assume there are at most 5 causal variables, i.e. set `L=5`. We set SuSiE prior variance to 0.1. The `susieR` function gives coefficients in standardized X, y scale:

```{r}
fitted_z <- susie_z(z = z_scores,R = R,L = 5)
```

We can check the 95% confidence set:
```{r}
summary(fitted_z)$cs
```
and the PIP plot
```{r}
susie_plot(fitted_z, y="PIP", b=b)
```

## Session information

Here are some details about the computing environment, including the
versions of R, and the R packages, used to generate these results.

```{r}
sessionInfo()
```
