---
title: "test.Rmd"
author: "Matthew Stephens"
date: "4/14/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# simulate data

This is Lei's example
```{r}
set.seed(777)
library(susieR)
X <- matrix(rnorm(1010 * 1000), 1010, 1000)
beta <- rep(0, 1000)
beta[1 : 200] <- 100
y <- X %*% beta + rnorm(1010)
s = susie(X,y,L=200)

plot(coef(s),beta)
s$sigma2

# fit <- lm(y ~ X - 1)
# mlr.p <- log(summary(fit)$coefficients[, 4])
# 
mar.p <- c()
mar.betahat = c()
for (i in 1 : 1000) {
 fit <- lm(y ~ X[, i] - 1)
  mar.p[i] <- log(summary(fit)$coefficients[, 4])
  mar.betahat[i] <- summary(fit)$coefficients[, 1]
}
# 
# pdf("pvalue.pdf", width = 10, height = 5)
# par(mfrow = c(1, 2))
# plot(mlr.p, ylab = "log(p-value)", main = "Multiple Linear Regression")
# abline(h = log(0.05 / 1000), lty = 2, col = "red")
# legend("right", lty = 2, col = "red", "log(0.05/p)")
# 
# plot(mar.p, ylab = "log(p-value)", main = "One-on-One Linear Regression")
# abline(h = log(0.05 / 1000), lty = 2, col = "red")
```

Notice that the coefficients are monotonic with betahat. Some shrinkage of zero values is evident, but it is not enough... presumably because sigma2 is way over-estimated. And further we see excessive shrinkage of true signals, presumably because sa2 is too small.
```{r}
plot(coef(s),mar.betahat)
```


Here we try fixing sigma2 to true value.
```{r}
s2 = susie(X,y,L=300,sigma2=1,estimate_sigma2=FALSE)
plot(coef(s2),beta)
```
it works!!

```{r}
plot(s$alpha[1,])
plot(s$alpha[2,])
```

Try with larger prior on effect sizes
```{r}
s3 = susie(X,y,L=200, sa2 = 100)
```


