---
title: "A minimal example"
author: "Matthew Stephens"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{minimal example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> This inherited vignette describes the **additive** interfaces retained
> in susieSlide. The original individual-data function is now
> `susie_additive()`. For the slider model use `susieSlide::susie()` and
> see the [slider introduction](slider-model.html).


```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,comment = "#",fig.width = 4.5,
                      fig.height = 3,fig.align = "center",
                      fig.cap = "&nbsp;",dpi = 120)
```

In this short vignette, we fit a sparse linear regression model with
up to $L > 0$ non-zero effects. Generally, there is no harm in
over-stating $L$ (that is, the method is pretty robust to overfitting),
except that computation will grow as $L$ grows.

Here is a minimal example:

```{r}
library(susieSlide)
set.seed(1)
n    <- 1000
p    <- 1000
beta <- rep(0,p)
beta[c(1,2,300,400)] <- 1
X   <- matrix(rnorm(n*p),nrow=n,ncol=p)
y   <- X %*% beta + rnorm(n)
res <- susie_additive(X,y,L=10)
plot(coef(res)[-1],pch = 20)
```

Plot the ground-truth outcomes vs. the predicted outcomes:

```{r fig.height=3.5, fig.width=3.5}
plot(y,predict(res),pch = 20)
```

## Session information

Here are some details about the computing environment, including the
versions of R, and the R packages, used to generate these results.

```{r}
sessionInfo()
```
